


```{r , include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,cache=TRUE)
library(knitr)
library(kableExtra)
```

# Penalty Kicks and MSNE

```{r}
library(tidyverse)

# Source: kaggle.com/mauryashubham/english-premier-league-penalty-dataset-201617
EPL<-(read.csv("penalty_data.csv") 
      %>% tibble()
      %>% mutate(Y = 1*(Scored=="Scored"))
        )
EPL %>% sample_n(10) %>% knitr::kable()
print(dim(EPL))
```


```{r}

EPLcounts<-matrix(0,3,3)
colnames(EPLcounts)<-c("L","C","R")
rownames(EPLcounts)<-c("L","C","R")
EPLscored<-EPLcounts

for (ii in c("L","C","R")) {
  for (jj in c("L","C","R")) {
    EPLcounts[ii,jj]<-dim(EPL %>% filter(Kick_Direction==ii & Keeper_Direction==jj))[1]
    EPLscored[ii,jj]<-dim(EPL %>% filter(Kick_Direction==ii & Keeper_Direction==jj & Scored=="Scored"))[1]
  }
} 

EPLKickerPayoff<-EPLscored/EPLcounts
colnames(EPLKickerPayoff)<-c("l","c","r")
print(EPLKickerPayoff)
```

## Other datasets:

> Azar, Ofer H., and Michael Bar-Eli. "Do soccer players play the mixed-strategy Nash equilibrium?." Applied Economics 43, no. 25 (2011): 3591-3601.

**Their Table 1, 3, 4**

```{r}
# Counts
ABcounts<-rbind(c(54,1,37),c(41,10,31),c(46,7,59))
#Stopped kicks
ABstopped<-rbind(c(16,0,0),c(4,6,1),c(0,0,15))

ABKickerPayoff<-1-ABstopped/ABcounts
colnames(ABKickerPayoff)<-c("l","c","r")
rownames(ABKickerPayoff)<-c("L","C","R")
print(ABKickerPayoff)
```

> Chiappori, P-A., Steven Levitt, and Timothy Groseclose. "Testing mixed-strategy equilibria when players are heterogeneous: The case of penalty kicks in soccer." American Economic Review 92, no. 4 (2002): 1138-1151.

**Their tables 3 and 4**

```{r}
# Counts
CLGcounts<-t(rbind(c(117,48,95),c(4,3,4),c(85,28,75)))


CLGKickerPayoff<-t(rbind(c(0.632,0.812,0.895),c(1.00,0,1.00),c(0.941,0.893,0.440)))
colnames(CLGKickerPayoff)<-c("l","c","r")
rownames(CLGKickerPayoff)<-c("L","C","R")
print(CLGKickerPayoff)
```

## MSNE - empirical payoffs

Note that we can solve for MSNE by noting that all expected payoffs must be constant, and that probabilities must add to 1:

$$
\begin{aligned}
EU(a)&=\text{a constant}\\
U_ip_i&=c\\
p_i&=U_i^{-1}c\\
\text{probabilities must add to 1:}
\\
p_i&=U_i^{-1}c / \sum_a U_i^{-1}c = U_i^{-1}c / \sum_a U_i^{-1}c
\end{aligned}
$$
Only works if payoffs are all non-negative, but we can always add a large constant to $U$ to fix this, and not change any of the incentives.
```{r}

U<-list(CLG=CLGKickerPayoff,AB=ABKickerPayoff,EPL=EPLKickerPayoff)

COUNTS<-list(CLG=CLGcounts,AB=ABcounts,EPL=EPLcounts)

D<-tibble()

action<-c("L","C","R")

for (ii in 1:length(U)) {
  uu<-U[[ii]]
  cc<-COUNTS[[ii]]
  
  pERow<-cc%*%c(1,1,1)/sum(cc)
  pECol<-t(cc)%*%c(1,1,1)/sum(cc)
  pNashCol<-solve(uu)%*%c(1,1,1)/sum(solve(uu)%*%c(1,1,1))
  pNashRow<-solve(t(1-uu))%*%c(1,1,1)/sum(solve(t(1-uu))%*%c(1,1,1))
  D<-tibble(pNashRow,pERow,pECol,pNashCol,action)%>% mutate(
    data = names(U)[ii]
    ) %>% rbind(D)
  
}

D %>% kbl()  %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")

```

```{r}
(
  ggplot()
  +geom_point(data=D,aes(x=pNashRow,y=pERow,color="Kicker",shape=data))
    +geom_point(data=D,aes(x=pNashCol,y=pECol,color="Keeper",shape=data))
  +geom_abline(slope=1,intercept=0)
  +geom_smooth(data=D,aes(x=pNashRow,y=pERow,color="Kicker"),method="lm",alpha=0.2)
    +geom_smooth(data=D,aes(x=pNashCol,y=pECol,color="Keeper"),method="lm",alpha=0.2)
  +theme_bw()
  +xlab("MSNE prediction")
  +ylab("Empirical frequencies")
)
```
