---
title: "ECON 4/6260 - Behaviroal Economics"
subtitle: "Introduction to Game Theory"
author: "Dr. James Bland"


output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r, include=FALSE}
library(tidyverse)
library(kableExtra)
knitr::opts_chunk$set(echo = FALSE,cache=FALSE)
```


# Introduction to Game Theory

## Game theory {.unlisted .unnumbered}

* **Game theory** is the analysis of strategic interactions.

* A **game** is a decision problem in which the final
outcome depends upon the decisions of all players, as
well as the state of the world.

* A **strategy** is a complete plan of action that describes
what a player will do under all possible circumstances.

* The **payoff matrix** represents the payoffs associated
with each possible combination of strategies.
    + Describe how every combination of action that each player could take results in payoffs for each player
    + Typically, we think about payoffs as utility , not money. Think about what that means about risk preferences. However in experiments, we pay in cash, not utility.
    
## Examples of payoff matrices {.unlisted .unnumbered}

:::: {style="display: flex;"}

::: {}

```{r}

# From Agner Table 1.4
Game1<-rbind(c("2, 2", "0, 0"),c("0, 0", "1, 1"))
colnames(Game1)<-c("left","right")
rownames(Game1)<-c("Up","Down")
Game1 %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic") %>% add_header_above(header = c("Game 1" = 3))

Game2<-rbind(c("5, 1", "2, 0"),c("5, 1", "1, 2"))
colnames(Game2)<-c("left","right")
rownames(Game2)<-c("Up","Down")
Game2 %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")%>% add_header_above(header = c("Game 2" = 3))

Game<-rbind(c("5, 5", "0, 6"),c("6, 0", "2, 2"))
colnames(Game)<-c("left","right")
rownames(Game)<-c("Up","Down")
Game %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")%>% add_header_above(header = c("Game 3" = 3))
```
:::

::: {}
$\quad$
:::

::: {}

```{r}


Game<-rbind(c("5, 5", "0, 2"),c("2, 0", "2, 2"))
colnames(Game)<-c("left","right")
rownames(Game)<-c("Up","Down")
Game %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")%>% add_header_above(header = c("Game 4" = 3))

Game<-rbind(c("2, 1", "0, 0"),c("0, 0", "1, 2"))
colnames(Game)<-c("left","right")
rownames(Game)<-c("Up","Down")
Game %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")  %>% add_header_above(header = c("Game 5" = 3))

Game3<-rbind(c("6, 2", "5, 1","4, 3"),c("3, 6", "8, 4","2, 1"),c("2, 8","9, 6", "3, 0"))
colnames(Game3)<-c("left","middle","right")
rownames(Game3)<-c("Up","Middle","Down")
Game3 %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")  %>% add_header_above(header = c("Game 6" = 4))
```

:::

::: {}
$\quad$
:::

::: {}

```{r}

# From Agner Table 1.4
Game1<-rbind(c("5, 2", "1, 1"),c("1, 1", "2, 5"))
colnames(Game1)<-c("left","right")
rownames(Game1)<-c("Up","Down")
Game1 %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic") %>% add_header_above(header = c("Game 7" = 3))

Game2<-rbind(c("4, 1", "2, 0"),c("5, 1", "1, 2"))
colnames(Game2)<-c("left","right")
rownames(Game2)<-c("Up","Down")
Game2 %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")%>% add_header_above(header = c("Game 8" = 3))

Game<-rbind(c("1, 1", "0, 0"),c("0, 0", "0, 0"))
colnames(Game)<-c("left","right")
rownames(Game)<-c("Up","Down")
Game %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")%>% add_header_above(header = c("Game 9" = 3))
```

:::

::::

## Best response {.unlisted .unnumbered}

A player’s *best response* correspondence describes the set of
strategies that maximize their utility, conditional on their opponent’s
strategy.

* Correspondence rather than function because sometimes there is
more than one thing that can maximize your utility

Examples

* Pure coordination

* Matching pennies

* Rock, paper, scissors (assume draw yields zero payoff)

* Rock, paper, scissors. Players get twice the payoff if they win by
playing rock.

## Dominant and dominated strategies {.unlisted .unnumbered}

A *dominant strategy* is a strategy that is a best response no matter
what your opponent’s strategy is

* *Strictly dominant strategy*: a strategy that you would strictly prefer to
play for any possible strategy your opponent could play

* *Weakly dominant strategy*: a strategy that does at least as well as any
other strategy you could play, against any possible strategy your
opponent could play

 A *Strictly dominated strategy* is a strategy that does strictly worse
against all possible opposing strategy than some other strategy that
you could play

## Elimination of strictly dominated strategies {.unlisted .unnumbered}


One solution concept for games is iterated elimination of strictly dominated strategies:

1. Rule out all strictly dominated strategies

2. Restrict analysis to the game where players’ strictly dominated strategies are eliminated

3. Go back to step 1

**Example: Cournot competition** -- Oligopoly, firms compete on quantity

Demand curve:

$$
P=a-bQ
$$
Identical firms, each has marginal cost $c$, no other costs

* $n=2$
* $n=3$ (or larger)


## Nash Equilibrium  {.unlisted .unnumbered}

* Nash equilibrium is a strategy profile such that each player is playing their best strategy given the decision of the other players.

* Nash’s Theorem states that every finite game that
is, every game in which all players have a finite number of pure strategies has a Nash equilibrium.

**Go back to examples earlier**

## Not all games have a Nash equilibrium in **pure** strategies {.unlisted .unnumbered}

**Example:** Rock, paper, scissors

**Mixed strategy Nash equilibrium** (MSNE)

Sometimes the comparative statics of MSNE are counter-intuitive, and sometimes not:

:::: {style="display: flex;"}

::: {}

```{r}
Game<-rbind(c("a, a", "0, 0"),c("0, 0", "1, 1"))
colnames(Game)<-c("left","right")
rownames(Game)<-c("Up","Down")
Game %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")
```

:::

::: {}
$\quad\quad$
:::

::: {}

```{r}
Game<-rbind(c("a, 1", "0, 0"),c("0, 0", "1, a"))
colnames(Game)<-c("left","right")
rownames(Game)<-c("Up","Down")
Game %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")
```

:::

::: {}
$\quad\quad$
:::


::: {}

```{r}
Game<-rbind(c("0, 0", "a, 1"),c("1, a", "1, 1"))
colnames(Game)<-c("left","right")
rownames(Game)<-c("Up","Down")
Game %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")
```

:::

::: {}
$\quad\quad$
:::

::: {}

```{r}
Game<-rbind(c("0, 0", "a, 1"),c("1, a", "1, 1"))
colnames(Game)<-c("left","right")
rownames(Game)<-c("Up","Down")
Game %>% kbl() %>% kable_styling(full_width=FALSE,htmltable_class="lightable-classic")
```

:::

::::


$$\text{how does MSNE change for } a>1 \text{?}$$

## Mixed strategies example: penalty kicks in soccer {.unlisted .unnumbered}


