---
title: "ECON 4/6260 - Behaviroal Economics"
subtitle: "01 - Introduction "
author: "Dr. James Bland"


output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setupIntro, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
set.seed(42)
```
# Introduction

## Boring stuff first: Expectation for class

* We will start and finish on time
* Attendance is expected
* All due dates are final outside of excused absences
* You are checking your UToledo email
* You have read and understand the Syllabus
* For all classroom experiments, at least one person will be paid. You will understand the payment mechanism before starting an experiment.

## What will we be studying?

* Insights from Psychology
  + Especially judgment & decision-making
* Extensions of "standard" (non-behavioral) economic theory
  + Markets
  + Choice theory
  + Games
* Experimental methods
* Non-experimental methods

## Doing economic research without experiments

Research question $\rightarrow$ I need data

* Data might not exactly answer the research question
* Often difficult to establish a causal relationship 
* Often requires assumptions to be made about the data that you cannot test.


I have data $\rightarrow$ what questions can I answer?

* Limits the questions you can answer
* Still difficult to establish a causal relationship
* Still (often) requires untestable assumptions

## Economic experiments

We will be using economic experiments to learn about concepts in Behavioral Economics.

Mostly, we will use *laboratory experiments*

* Use people who have volunteered to participate. Usually undergrads.
  + In this class: you
* Give them a task to perform, and observe their behavior.
* Pay them money based on their decisions

## Why pay money?

Mankiw's *10 Principles of Economics*: "4. People respond to incentives"

If I pay you enough, I will have a good idea about what your incentives are. 

This means that:

* Participants need to understand {\bf how} decisions determine payments
  + Explain to participants how this works before they make decisions (and maybe verify)
* Participants need to believe that decisions *will* determine payments 
  + The experimenter will never lie to participants
* Payments need to care enough about the reward mechanism that they will respond to it
  + We can't pay you in chocolate or Monopoly money, or ask you to make hypothetical choices. We use US dollars!

## How this will work for ECON4/6260


* I have approval from the Department of Economics to pay you in US dollars
* If I pay you, I will need to record your name and how much I paid you
* I don't have enough to pay everybody every time. I will randomly select (at least) one person to be paid after each experiment. You will not know who is getting paid until all decisions have been made.
* As this is an online class, I will give you several choices of how you can be paid

* This is not my money, so don't feel badly about taking it
* Don't rely on it for lunch, car payments, etc.

## And now for our first experiment

...

## Reasons for running an experiment 

Studying theoretical predictions

* Point and comparative static predictions
* Comparing the predictions of two or more theories

Studying behavioral assumptions

* Maybe a theory performs poorly. Can we work out which assumption(s) is (are) causing this?
* E.g. Are people selfish? Do people maximize expected value? Do people ignore sunk costs?



Establishing behavioral regularities

* Does the theory make more than one prediction? (multiple equilibria)
* Explore behavior to help us make better behavioral assumptions


Further reading: 

> Nikiforakis, Nikos. ["For the student: experimental economics."](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1467-8462.2010.00607.x?casa_token=K6AYdLRDvdAAAAAA%3AideO69vzWNyKiBui76sE4jOLASAPBw0JoDZ97Uxy0POWYZ_pOV-T1feeFWAOtUdY9fWIYXcFSNGe3RJ6) Australian Economic Review 43, no. 3 (2010): 337-345.

## Studying theoretical predictions - I

Suppose that you have a theory that makes a prediction $f(x)$ for situation $x$.

e.g.:

*  When the temperature is $x^\circ\mathrm F$, $f(x)=10+3x$ pints of ice cream will be sold

**Point predictions**: Given situation $x$, $f(x)$ will happen

* When the temperature is $90^\circ \mathrm F$, $280 = 10 + 3\times 90$ pints of ice cream will be sold

**Comparative static predictions**: When $x$ increases, $f(x)$ increases (or decreases, or nothing happens to $f(x)$)

* When temperature increases, ice cream sales increase. 
* When temperature increases by $1^\circ \mathrm F$, ice cream sales increase by 3 pints.

*When are each of these useful?*

## Studying theoretical predictions - II

**Comparing two or more theories**

I have two theories about how $x$ affects an outcome. These theories are $f(x)$ and $g(x)$. I wish to understand whether one makes better predictions than another.

```{r Intro-comparing-theories,fig.width=10,fig_height=8}
x<-seq(0,1,length=101)
y<-x*(1-x)
pltData<-data.frame(y,x)
pltData$Theory <- "f(x)"
tmp<-pltData
tmp$y<-x
tmp$Theory<-"g(x)"
pltData<-rbind(pltData,tmp)

xdata<-runif(n=1000)
ydata<-1/(1+exp(-((log(xdata)-log((1-xdata)))+0.2*rnorm(n=1000))))
dataXY<-data.frame(xdata,ydata)

plt<-(
  ggplot(pltData,aes(x=x,y=y,color=Theory))
  +geom_line(size=2)
  +geom_point(data=dataXY,aes(x=xdata,y=ydata),size=0.5,color="black")
  +theme_bw()
  +xlab("x")
  +ylab("")
)
plt
```

## Studying theoretical predictions - III

**Comparing two or more theories**: Sometimes both are important!

```{r Intro-comparing-theories-III,fig.width=10,fig_height=8}

xdata<-runif(n=1000)
group<-(runif(n=1000)<0.7)
yprediction<-xdata*(1-xdata)
yprediction[group]<-xdata[group]

ydata<-  1/(1+exp(-((log(yprediction)-log((1-yprediction)))+0.4*rnorm(n=1000))))
dataXY<-data.frame(xdata,ydata)

plt<-(
  ggplot(pltData,aes(x=x,y=y,color=Theory))
  +geom_line(size=2)
  +geom_point(data=dataXY,aes(x=xdata,y=ydata),size=0.5,color="black")
  +theme_bw()
  +xlab("x")
  +ylab("")
)
plt
```

## Studying theoretical predictions - IV

**Comparing two or more theories**: Sometimes they are all useless

```{r Intro-comparing-theories-IV,fig.width=10,fig_height=8}

xdata<-runif(n=1000)

yprediction<-1-xdata

ydata<-  1/(1+exp(-((log(yprediction)-log((1-yprediction)))+0.4*rnorm(n=1000))))
dataXY<-data.frame(xdata,ydata)

plt<-(
  ggplot(pltData,aes(x=x,y=y,color=Theory))
  +geom_line(size=2)
  +geom_point(data=dataXY,aes(x=xdata,y=ydata),size=0.5,color="black")
  +theme_bw()
  +xlab("x")
  +ylab("")
)
plt
```

## Economic experiments and causality

Conduct a study that resembles a drug trial.

Often referred to as a "randomized controlled trial"


Drug trials:

* Randomly assign people to two groups
* One group gets the treatment (i.e. they get the drug)
* One group doesn't get the drug (or a placebo)
* Random assignment ensures that differences between groups' health outcomes could only be due to the drug

Economic experiments:

* Instead of health outcomes, we are interested in peoples' decisions
* Instead of drugs, we will be treating people with *different incentives*
* Many field experiments use this format, e.g the work associated with the [2019 Nobel prize in Economics](https://www.nobelprize.org/prizes/economic-sciences/2019/press-release/)

## Complement to data from outside the lab

Example: ECON1200 model of a market.

* Buyers with values
* Sellers with costs
* If we know all of these, we can calculate the equilibrium price and quantity


Outside the lab:

* Buyers might not want to tell you their values
* Sellers will probably lie to you about costs
* We can easily observe price and quantity, but not the values and costs that got us there


Inside the lab:

* We can assign buyers and sellers values and costs
* Hence, we know what the theory predicts, and we can observe the outcomes.

## A roadmap for this course

Usually, I will introduce a topic with an experiment.

Then, you will have a homework assignment based on the data from the experiment

We will discuss this in the next class

For 6260 students, we will use our extra hour mostly as a reading group. I will assign the first few readings, but I am very open to suggestions. 

## Outline of topics covered (not necessarily in order)

* Game theory
  + How do economists model interactions in small groups?
  + Key concepts: Nash Equilibrium, best response, mixed and pure strategies
  + Single-shot games, sequential games, Bayesian games
* Behavioral Game Theory
  + Incorporating insights from decision theory into game theory
  + Noisy behavior in equilibrium
  + Other departures from the *Homo Economicus* assumption
  + Non-standard preferences
* Preferences & Decisions
  + Risk
  + Probability judgment
  + Other-regarding preferences
  

  







