---
title: "ECON 4/6260 - Behaviroal Economics"
subtitle: "02 - Markets"
author: "Dr. James Bland"


output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---



```{r setupMarkets, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
set.seed(42)
```

## Markets

### Markets


* Probably one of the first things you learned about in a *Principles of Microeconomics* course

* A group of buyers, each with values for purchasing items
  + Implies a downward-sloping demand curve
* A group of sellers, each with costs of selling item
  + Implies an upward-sloping supply curve

* This is enough to calculate the equilibrium price and quantity

However:
* One cannot just ask buyers what there values are
* One cannot just ask sellers what their values are

(why not?)

---

### An experiment

...

---

### Induced value theory

Simple version: *Pay people enough money, and they will behave as if these are their incentives*

More formally, if for a participant in an experiment:
* taking action $a$ results in payment $A
* taking action $b$ results in payment $B

Then they will only choose action $a$ if $A \geq B$, and will only choose action $b$ if $B\geq A$.

**Caveats**: $A$ and $B$ need to be large enough that participants pay attention to them (salience)

* Don't pay in Monopoly money, chocolate, etc
* Maintain assumption that participants don't care about the earnings of others (selfishness)
* Participants need to understand how their actions lead to payments
* Participants need to believe that their actions will lead to payments

---

### In market experiments

* Each buyer $i$ has values $v_{i,1}\geq v_{i,2}\geq v_{i,3} \ldots$.
* Each seller $j$ has costs $c_{i,1}\leq c_{i,2}\leq c_{i,3} \ldots$.
* If buyer $i$ buys from seller $j$ at price $p$:
  + Buyer $i$ earns $\$ v_{i,1}-p$
  + Seller $j$ earns $\$ p-c_{j,1}$
  
(if this is the first trade for both)

---

### Numerical example - Buyers



There are 10 buyers, each buyer has a value equal to a 15-sided die roll plus 3, and can buy at most 3 items:
```{r,echo=T}
set.seed(42)
nBuyers<-10
nItems<-3
Demand<-data.frame()
for (bb in 1:nBuyers) {
  # value for that individual buyer
  v<-sample(1:15,nItems,replace=T)+3
  tmp<-data.frame(v)
  tmp$BuyerID<-bb
  Demand<-rbind(Demand,tmp)
}
plt<-(
  ggplot(data=Demand,aes(x=v))+geom_histogram(bins=10)
  +theme_bw()
  +xlab("Values")
  +ylab("count")
)

```


---

```{r, echo=F,fig.width=6}
print(plt)
```


---

### Numerical example - Sellers


There are 20 sellers, each seller has a cost equal to a 10-sided die roll, and can sell at most 2 items:
```{r,echo=T}
nSellers<-10
nItems<-2
Supply<-data.frame()
for (ss in 1:nSellers) {
  # value for that individual buyer
  c<-sample(1:10,nItems,replace=T)
  tmp<-data.frame(c)
  tmp$SellerID<-ss
  Supply<-rbind(Supply,tmp)
}
plt<-(
  ggplot(data=Supply,aes(x=c))+geom_histogram(bins=10)
  +theme_bw()
  +xlab("Costs")
  +ylab("count")
)

```

---

```{r, echo=F,fig.width=6}
print(plt)
```



---

### Numerical example - Supply and demand curves

```{r, echo=T}
P<-seq(0,20,length=1000) # prices to evaluate quantity
SF<-function(p) {
  sum(Supply$c<=p)
}
DF<-function(p) {
  sum(Demand$v>=p)
}

SupplyDemand<-data.frame()
for (pp in P) {
  p<-pp
  Q<-DF(pp)
  type="demand"
  SupplyDemand<-rbind(SupplyDemand,data.frame(p,Q,type))
  Q<-SF(pp)
  type="supply"
  SupplyDemand<-rbind(SupplyDemand,data.frame(p,Q,type))
  
}
plt<-(ggplot(data=SupplyDemand,aes(x=Q,y=p,color=type))
  +geom_path()+theme_bw()
  +xlab("Q")+ylab("P")
  +theme(legend.title = element_blank(),
         legend.position=c(0.8,0.8))
)
```
---

```{r, echo=F,fig.width=4}
print(plt)
```

---



### Point predictions from markets


* Equilibrium price and quantity 
* Which buyers trade?
* Which sellers trade?
* Consumer and producer surplus 

```{r, echo=T}
ExcessDemand<-apply(as.matrix(P),
                    FUN = function(x) {DF(x)-SF(x)},
                    MARGIN=1)

Peq<-P[which.min(abs(ExcessDemand))]
Qeq<-c(DF(Peq),SF(Peq))

# equilibrium price
print(Peq)

# equilibrium quantity
print(Qeq)

```



---

### Point predictions



```{r,echo=T}
plt<-(plt
      +geom_hline(yintercept=Peq,
                  color="black",
                  linetype="dashed")
      +geom_vline(xintercept=Qeq,
                  color="black",
                  linetype="dashed")
)

CS<-sum((Demand %>% filter(v>=Peq))$v)
print(CS)

PS<-sum((Supply %>% filter(c<=Peq))$c)
print(PS)


```
 
--- 
 
```{r, echo=F,fig.width=6}
plt
```


\

---

### Comparative static predictions

What happens to any of the above if?

* Supply or demand shift?
* The government imposes a tax?
* The government subsidizes the good?
* The government imposes a price floor or ceiling?
  
---
  
### Other take-aways

* Market theory remains silent on *how* buyers and sellers discover the equilibrium price
* Market theory remains silent on whether different trading mechanisms affect price discovery.






