---
title: "ECON 4/6260 - Behaviroal Economics"
subtitle: "03 - Risk and Decision Making"
author: "Dr. James Bland"


output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

<style>
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>

```{r setupRisk, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,cache=T)
library(ggplot2)
library(dplyr)
set.seed(42)
```

# Risk and decision-making


## Risk and uncertainty

Common attributes of many decisions that we make:

* Should I drive my car today?
* Should eat something that has a "sell by" date of one week ago?
* If I go to the grocery store, what is the probability that I will contract COVID-19?
* Should I study and revise a particular slide for an exam?

Here we will make a distinction between:

* **Risk** I know (or at least can estimate) the probabilities of events when I make my decisions:
  + Die rolls
  + Gambling at a casino
  + Buying car insurance (if I know the probability that I will crash)
* **Uncertainty** I don't know the probabilities when I make my decisions:
  + If you don't know the motivations of others
  + If the decision problem is too complicated
  
In this chapter, we will focus on **Risk**

## Example: Insurance

> I can buy car insurance for $P. The following table shows probabilities and outcomes of particular events.

```{r, echo=F}

W<-30000

Event<-c("1 - no crash","2 - a minor ding","3 - panel replaced","4 - three panels replaced","5 - new engine","6 - car totaled")
Probability<-rep(paste("1/6"),6)
Loss<-(0:5)^(3)*100
Initial.Wealth<-rep(W,6)
Without.Insurance<-Initial.Wealth-Loss
With.Insurance<-paste(Initial.Wealth," - P")



Tab<-data.frame(Event,Probability,Loss,Initial.Wealth,Without.Insurance,With.Insurance)

knitr::kable(Tab)

```

How much am I willing to pay for this insurance?


## Expected value

One approach would be to determine the expected value of having and not having insurance. The insurance guarantees that my bank balance will be $\$ `r paste(W)`-P$. What is my "average" bank balance if I do not purchase the insurance?

**Definition: Expected value **: if $X$ is a real-valued random variable with probability mass function $p_X(x)=\Pr(X=x)$, and support $\mathbb X$ (the possible values of $X$) then the expected value (or expectation) of $X$ is:

$$
\begin{aligned}
E(X)&=\sum_{x\in\mathbb X} x p_X(x)
\end{aligned}
$$
  
*example on doc cam*


## Expected Value

Support : $\mathbb X =\{ `r paste(Without.Insurance)1\}$

Probability mass function:
$$p_X(x)=\begin{cases}
\frac16 &\text{if }x\in\mathbb X\\
0&\text{Otherwise}
\end{cases}
$$

```{r,echo=F,results="asis"}
SumStr<-paste(Without.Insurance,"\\times",rep("\\frac16",6))

PrintThis<-c(
  "$$\\begin{aligned}",
  "E(x)&=\\sum_{x\\in\\mathbb X} x p_X(x) \\\\ \n",
  paste("&=",paste(SumStr,collapse="+"),collapse=""),
  "\\\\ \n",
  paste("&=",sum(Without.Insurance/6)),
  "\\\\ \n",
  "\\end{aligned}$$"
)


cat(PrintThis)

```

Therefore you will be willing to pay $P=\$30000-`r paste(sum(Without.Insurance/6))`=\$`r paste(30000-sum(Without.Insurance/6))`$ for the insurance.



## But most people care about risk!

$$
\begin{aligned}
\mathbb X = \{\$1,\$2,\$3\}
,\ L_1:\ p^1_X = \{0.5,0.0, 0.5\} ,\quad L_2:\ p^2_X = \{0.0,1.0, 0.0\}
\end{aligned}
$$

Most people would choose $L_2$ in this situation.


## Expected utility theory

Instead of maximizing:

$$
E(X)=\sum_{x\in\mathbb X} xp_X(x)
$$

People maximize expected utility:
$$
EU=\sum_{x\in\mathbb X} u(x)p_X(x)
$$

where $u(x)$ is a "utility (of money) function"

* $u$ is *increasing*
* If $u$ is concave, this I am risk-averse
* If $u$ is convex, this I am risk-loving


## Insurance with risk aversion

Assume $u(x)=x^\alpha$ (replace $\alpha=0.5$ for $u(x)=\sqrt x$ if it helps)

Expected utility without insurance

```{r,results="asis"}
SumStr<-paste(Without.Insurance,"^\\alpha")

PrintThis<-c(
  "$$\\begin{aligned}",
  "E(u(x))&=\\sum_{x\\in\\mathbb X} u(x) p_X(x) =\\sum_{x\\in\\mathbb X} x^\\alpha p_X(x) \\\\ \n",
  paste("&=\\frac16\\left(",paste(SumStr,collapse="+"),"\\right)",collapse=""),
  "\\end{aligned}$$"
)
cat(PrintThis)

```
For $\alpha=0.5$, this becomes `r paste(round(sum(Without.Insurance^0.5 /6),1))`

Expected utility with insurance:

$$
E(u(30000-P))=(30000-P)^\alpha
$$

So you should buy insurance if and only if:

```{r,results="asis"}
SumStr<-paste(Without.Insurance,"^\\alpha")

PrintThis<-c(
  "$$\\begin{aligned}",
  "(30000-P)^\\alpha &\\geq",
  paste("\\frac16\\left(",paste(SumStr,collapse="+"),"\\right)",collapse=""),
  "\\\\",
  paste("P&\\leq 30000-\\left(\\frac16\\left(",paste(SumStr,collapse="+"),"\\right)\\right)^{\\frac1\\alpha}",collapse=""),
  "\\end{aligned}$$"
)
cat(PrintThis)

```


## Insurance with risk aversion - graphically

```{r, echo=F}

r<-c(1,seq(0.2,1.2,length=1001))

p<-c()
for (rr in r) {
  p=c(p,max(c(W-sum((1/6)*Without.Insurance^rr),0)))
}
plotThis<-data.frame(r,p)

(
  ggplot(data=plotThis,aes(x=r,y=p))
  +geom_line(size=2)
  +theme_bw()
  +xlab("risk aversion parameter r")
  +ylab("Willingness to pay ($)")
  +geom_vline(xintercept=1,linetype="dashed")
)





```
